---
title: "In-class Exercise 4"
author: "Meng Tingyu"
date: "September 18, 2026"
date-modified: today
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

## Getting Started

```{r}
pacman::p_load(spdep, tmap, ggstatsplot, tidyverse, knitr, GWmodel, sf)
```

## Preparing the Data

### Importing Hunan shapefile

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

### Importing Hunan_2012 table

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Joining Hunan and Hunan_2012

```{r}
hunan_sf <- left_join(hunan_sf, hunan2012) %>%
  select(1:3, 7, 15, 16, 31, 32)
```

## Mapping GDPPC

```{r}
basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5)

gdppc <- qtm(hunan_sf, "GDPPC") +
  tm_layout(legend.position = c("left", "bottom"))

tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

## Converting to SpatialPolygonDataFrame

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

## Determining Adaptive Bandwidth

### Cross-validation

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

### AICc

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                 data = hunan_sp,
                 approach = "AIC",
                 adaptive = TRUE,
                 kernel = "bisquare",
                 longlat = T)
```

## Computing Geographically Weighted Summary Statistics

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)
```

## Preparing the Output Data

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

## Visualising Geographically Weighted Summary Statistics

```{r}
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Local Mean")) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Distribution of Geographically Weighted Mean") +
  tm_layout(legend.position = c("left", "bottom"))
```
