---
title: "Hands-on Exercise 01: Geospatial Data Wrangling"
author: "Meng Tingyu"
date: "last-modified"
date-format: "DD MMM YYYY"
---

## Overview

Placeholder. I will replace this with the actual write-up after the first lesson.

The plan for this exercise is to import geospatial data in shapefile and KML format, import an aspatial dataset and convert it into a spatial object, and check that everything ends up in a consistent coordinate reference system before any analysis begins.

## Getting started

```{r}
pacman::p_load(sf, tidyverse, tmap)
```

## Importing geospatial data

```{r}
#| eval: false
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

## Checking the coordinate reference system

Always confirm the CRS before doing anything else. Singapore data arrives in a mix of SVY21 and WGS84, and any calculation involving distance or area needs a projected CRS.

```{r}
#| eval: false
st_crs(mpsz)
```

```{r}
#| eval: false
mpsz3414 <- st_transform(mpsz, crs = 3414)
```

::: {.callout-note}
## Note to self
`st_set_crs()` and `st_transform()` do different things. `st_set_crs()` only relabels the data as being in a given CRS without moving any coordinates, so it is the right tool when the CRS tag is wrong but the numbers are fine. `st_transform()` actually reprojects the coordinates. Using the first when I meant the second silently produces wrong distances.
:::
