---
title: "Hands-on Ex06: Geographical Segmentation with Spatially Constrained Clustering"
subtitle: "Delineating ICT regions across the townships of Shan State, Myanmar"
author: "Meng Tingyu"
date: "October 2, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the code"
    code-tools: true
    fig-align: center
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

## What this exercise is about

This exercise groups the 55 townships of Shan State in Myanmar into regions with similar access to information and communication technology. The grouping uses six household ICT measures from the 2014 census, radio, television, land line phone, mobile phone, computer and internet at home.

Planners often want regions that are both homogeneous and joined together on the map. Ordinary cluster analysis only looks at the first condition. The second half of the exercise adds the map back in, first with SKATER, which cuts a minimum spanning tree of neighbouring townships, and then with ClustGeo, which lets me choose how much weight geography gets.

The work is split into six parts. Part 1 explores the variables with charts and choropleth maps, Part 2 checks correlations, and Part 3 runs hierarchical clustering. Part 4 and Part 5 cover the two spatially constrained methods, and Part 6 profiles the clusters.

## Setting up

### The data

Two datasets come from the Myanmar Information Management Unit (MIMU).

The first is `myanmar_township_boundaries`, an ESRI shapefile with the polygon boundary of every township in Myanmar. The second is `Shan-ICT.csv`, an extract of the 2014 Myanmar Population and Housing Census at township level. It has the number of households in each township and the number that own each of the six ICT items.

I put the shapefile in `data/geospatial` and the csv in `data/aspatial`, and added an empty `data/rds` folder for the intermediate files. The notes use `chap12/data/...`, and I changed every path to match this folder layout.

### Packages

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

The packages fall into five groups. **sf** and **spdep** handle the spatial data, and **tidyverse** covers readr, dplyr, tidyr and ggplot2 for the attribute data. **tmap** draws the choropleth maps. **corrplot**, **ggpubr**, **heatmaply**, **psych** and **GGally** are used to visualise the variables, and **cluster**, **factoextra**, **NbClust** and **ClustGeo** run the cluster analysis.

::: {.callout-note}
## rgdal is no longer needed

The package list in the notes mentions **rgdal**. It was retired from CRAN in 2023, and sf now does all of its reading and writing work, so I left it out. The `p_load()` call in the notes already does not include it.
:::

### Importing the township boundaries

`st_read()` reads the whole country. I keep only the three parts of Shan State and the six columns from state name to township code.

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The full layer has 330 townships and 15 fields in WGS 84. After filtering there are 55 townships, 24 in Shan (North), 21 in Shan (South) and 10 in Shan (East).

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```

`shan_sf` is an sf data frame, so it works with dplyr verbs and `glimpse()` like any tibble. All six attribute columns are character, and the geometry is MULTIPOLYGON in degrees.

### Importing the census extract

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

```{r}
summary(ict)
```

The table has 11 fields and 55 rows, one per township. Four fields are identifiers and seven are counts of households. Every count is strongly right skewed. Total households range from 3,318 to 82,604, and the mean of 18,369 sits well above the median of 13,685.

### Deriving penetration rates

The raw counts mostly measure how big a township is. A township with many households will also have many households with a radio, and that says nothing about how common radios are there. I divided each count by the total number of households and multiplied by 1,000 to get the number of owning households per thousand.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

Six new fields now sit at the end of the table, `RADIO_PR`, `TV_PR`, `LLPHONE_PR`, `MPHONE_PR`, `COMPUTER_PR` and `INTERNET_PR`. Television is the most common item, with a median of 517 households per thousand. Computers and home internet are rare, with medians of 19 and 23 per thousand.

## Part 1, Exploratory data analysis

### Distribution of radio ownership

A histogram shows the overall shape of a variable, and a boxplot picks out outliers. I started with the raw count of households with a radio.

```{r}
#| fig-width: 7
#| fig-height: 4
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
#| fig-width: 7
#| fig-height: 3
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

The count is heavily right skewed, with a skewness of about 2.6. Three quarters of the townships have fewer than 6,200 radio households, and three outliers sit far to the right, Taunggyi with 30,176, Lashio with 15,307 and Nyaungshwe with 13,801. These are also among the four townships with the most households.

Next I plotted the radio penetration rate.

```{r}
#| fig-width: 7
#| fig-height: 4
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
#| fig-width: 7
#| fig-height: 3
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

The course asks what the two distributions reveal. Dividing by households removes most of the skew. `RADIO_PR` has a mean of 216 and a median of 211 per thousand, and its skewness drops to 0.48. The boxplot now flags only one outlier, Ywangan, where 485 of every thousand households own a radio. Taunggyi, the largest outlier in the raw counts, has a rate of 365 and sits inside the whiskers, so its earlier position came from size alone.

### Distributions of all six rates

I drew one histogram for each rate and arranged them with `ggarrange()` from **ggpubr**.

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
#| fig-width: 11
#| fig-height: 6
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

The six rates fall into two shapes. Radio, television and mobile phone are roughly bell shaped, with television slightly skewed to the left (skewness -0.38). Land line phone, computer and internet are right skewed, with skewness between 1.3 and 1.8, because most townships have very low rates and a handful have much higher ones.

### Joining the census data to the boundaries

The choropleth maps need the rates attached to the polygons. `left_join()` keeps `shan_sf` as the base, so the result is still an sf object, and adds the columns of `ict_derived` by matching `TS_PCODE`.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

All 55 township codes match in both directions, so no township is dropped or left without data.

::: {.callout-note}
## Why TS.x and TS.y appear

The notes mention a message naming the join field. With `by` written out there is no message. Both tables have `DT`, `DT_PCODE` and `TS` columns besides the key, so dplyr keeps both copies and adds the suffixes `.x` and `.y`. I checked that `TS.x` and `TS.y` hold the same township name in all 55 rows. That is why the later code refers to `TS.x`.
:::

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

### Choropleth map of the radio penetration rate

`qtm()` gives a quick map with default settings.

```{r}
#| fig-width: 7
#| fig-height: 7
qtm(shan_sf, "RADIO_PR")
```

Radio ownership is highest in the west of Shan South and Shan North. Ywangan, Pindaya and Lawksawk near Taunggyi and Mabein in the north west all exceed 380 per thousand. The lowest rates are in the north east along the Chinese border, Konkyan and Laukkaing in the Kokang area at 21 and 53, and Narphan and Pangwaun in the Wa area at about 75.

### Counts versus rates on the map

To show how much the counts depend on township size, I mapped total households next to the number of radio households. The notes use tmap v3 syntax here, and I translated it to v4, where the classification goes into `tm_scale_intervals()`.

```{r}
#| fig-width: 12
#| fig-height: 7
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_polygons(fill = "TT_HOUSEHOLDS",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5),
              fill.legend = tm_legend(
                title = "Total households")) + 
  tm_borders(col_alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_polygons(fill = "RADIO",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5),
              fill.legend = tm_legend(
                title = "Number Radio")) + 
  tm_borders(col_alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

The two maps look almost the same. Taunggyi and Lashio are in the top class on both, and the correlation between total households and radio households across the 55 townships is 0.94.

Next I put total households next to the rate. The notes do this with two variables in one `tm_polygons()` call and `tm_facets(sync = TRUE)`. In v4, a vector of variables in `fill` already produces facets, and `fill.free = TRUE` gives each one its own Jenks classes.

```{r}
#| fig-width: 12
#| fig-height: 7
tm_shape(shan_sf) +
  tm_polygons(fill = c("TT_HOUSEHOLDS", "RADIO_PR"),
              fill.scale = tm_scale_intervals(
                style = "jenks"),
              fill.free = TRUE,
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"))) +
  tm_facets(ncol = 2) +
  tm_layout(outer.margins = 0, asp = 0)
```

::: {.callout-warning}
## Translating the tmap v3 code

Three changes were needed for tmap 4.4. `tm_fill(col = ..., n = 5, style = "jenks", title = ...)` became `tm_polygons(fill = ..., fill.scale = tm_scale_intervals(style = "jenks", n = 5), fill.legend = tm_legend(title = ...))`. `tm_borders(alpha = 0.5)` became `tm_borders(col_alpha = 0.5)`, since a border is a line. `tm_legend(legend.position = ...)` was replaced by a position inside the `fill.legend` argument.
:::

The course asks whether I can see the differences. The rate map breaks the link with size. Taunggyi, the largest township, drops to the fourth of five classes for `RADIO_PR`, and Lashio, the second largest, falls to the middle class at 236 per thousand. Small townships such as Mabein (8,632 households) and Monghpyak (6,165) move into the top class. The correlation between total households and `RADIO_PR` is only 0.34, which confirms that the rate measures something different from size.

## Part 2, Correlation analysis

Cluster analysis treats each variable as a separate dimension. Two variables that carry almost the same information would count that information twice, so I checked the correlations between the six rates first.

```{r}
#| fig-width: 7
#| fig-height: 7
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

`COMPUTER_PR` and `INTERNET_PR` have the strongest correlation, 0.87, so I keep only one of them. Following the notes, I keep `COMPUTER_PR`. The next highest pairs are `MPHONE_PR` with `COMPUTER_PR` at 0.74 and `TV_PR` with `MPHONE_PR` at 0.72. They are high but well below 0.87, and the notes keep all of them.

Radio stands apart from the rest. Its correlation with the other rates ranges from -0.31 with land line phones to 0.24 with television. Radio is common in townships where the more modern items are rare, so it adds information the other variables do not have.

## Part 3, Hierarchical cluster analysis

### Extracting the clustering variables

I dropped the geometry and kept the township name with the five clustering variables.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

`INTERNET_PR` is left out because of its correlation with `COMPUTER_PR`.

The clustering functions use row names as labels, so I replaced the row numbers with township names.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Then I dropped the `TS.x` column, since the name is now stored in the row names.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

::: {.callout-note}
## Saving shan_ict before reading it back

The notes read `shan_ict.rds` at this point but never write it. I added the `write_rds()` so the file exists before `read_rds()` loads it.
:::

```{r}
write_rds(shan_ict, "data/rds/shan_ict.rds")
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

### Data standardisation

The five variables have very different ranges. Mobile phones run from 36 to 735 per thousand, and computers from 3 to 92. Without standardisation, a variable with a large range has more pull on the distances between townships. The notes show two ways to standardise.

#### Min-max standardisation

`normalize()` from **heatmaply** rescales each variable to run from 0 to 1.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

Every variable now has a minimum of 0 and a maximum of 1. The means differ, from 0.24 for `COMPUTER_PR` to 0.54 for `TV_PR`, because the shape of each distribution is kept.

#### Z-score standardisation

`scale()` subtracts the mean and divides by the standard deviation. I used `describe()` from **psych** instead of `summary()` because it reports the standard deviation as well.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

All five variables now have a mean of 0 and a standard deviation of 1. The skewness values are the same as before standardisation, 1.80 for `COMPUTER_PR` and 1.37 for `LLPHONE_PR`.

::: {.callout-tip}
## When to use z-scores

The notes warn that z-scores should only be used if the variables can be assumed to come from a normal distribution. `COMPUTER_PR` and `LLPHONE_PR` are clearly skewed, so a z-score of 3 means something different for them than for `TV_PR`. Min-max scaling makes no such assumption, but a single extreme value squeezes everyone else into a narrow part of the 0 to 1 range.
:::

#### Visualising the standardised variables

I compared the histogram of `RADIO_PR` before standardisation and after each method.

```{r}
#| fig-width: 12
#| fig-height: 4
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

```{r}
#| fig-width: 12
#| fig-height: 4
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

The course asks what statistical conclusion the histograms support. The three panels have the same shape and differ only in the numbers on the x axis. Both methods are linear transformations, so they change the location and scale of a variable but not its skewness or the relative positions of the townships. Standardisation evens out the weight of each variable in the distance calculation, and it does not make a skewed variable normal.

### Computing the proximity matrix

`dist()` supports six distance measures, euclidean, maximum, manhattan, canberra, binary and minkowski, with euclidean as the default. I computed the euclidean distance between every pair of townships.

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

```{r}
#| eval: false
proxmat
```

The printed matrix has 1,485 distances, one for each pair of the 55 townships, so I left it out of the rendered page.

::: {.callout-warning}
## The distances use the unstandardised values

The notes standardise the variables and then build `proxmat` from `shan_ict`, the raw rates. I followed the notes so that the results stay comparable, but I checked how much each variable contributes.

```{r}
round(sapply(shan_ict, var) / sum(sapply(shan_ict, var)), 3)
```

`MPHONE_PR` and `TV_PR` together account for 81 percent of the total variance, and `COMPUTER_PR` for less than 1 percent. In the raw scale the clusters are mostly built from mobile phones and television. I come back to this at the end of Part 3.
:::

### Computing hierarchical clustering

`hclust()` supports eight agglomeration methods, ward.D, ward.D2, single, complete, average (UPGMA), mcquitty (WPGMA), median (WPGMC) and centroid (UPGMC). Following the notes, I used ward.D.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

```{r}
#| fig-width: 12
#| fig-height: 6
plot(hclust_ward, cex = 0.6)
```

The tree has one very tall final merge at a height of about 4,336, which joins a group of 31 townships with a group of 24. Below that, the next merges happen at 1,820 and 1,443.

### Selecting the optimal clustering algorithm

`agnes()` from **cluster** works like `hclust()` but also returns the agglomerative coefficient. It measures how strong the clustering structure is, and values closer to 1 mean a stronger structure. I computed it for four methods.

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

Ward's method has the highest coefficient at 0.943, ahead of complete linkage at 0.895, average at 0.813 and single linkage at 0.663. Single linkage chains townships one at a time, which gives the weakest structure. I use Ward's method from here on.

### Determining the optimal number of clusters

The notes list three common methods for choosing the number of clusters, the elbow method, the average silhouette method and the gap statistic, and use the gap statistic.

#### Gap statistic method

The gap statistic compares the within cluster variation for each k with the value expected from data with no cluster structure. The best k is the one with the largest gap. `clusGap()` from **cluster** computes it, and `hcut()` from **factoextra** does the hierarchical clustering inside each run.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

```{r}
#| fig-width: 8
#| fig-height: 5
fviz_gap_stat(gap_stat)
```

The largest gap is at k = 1 (0.274), so the statistic suggests the townships form one group. That is not useful for segmentation. Among k from 2 to 10, the largest gap is at k = 6 (0.2222), which is why the notes choose six clusters.

The choice of six is weak. The gap at k = 5 is 0.2216 and at k = 4 is 0.2184, both within 0.004 of k = 6, while the standard error for each is about 0.04. Any k from 4 to 6 fits the data about equally well. I keep six so that the rest of the results can be compared with the notes.

The notes also mention **NbClust**, which reports 30 indices for the number of clusters and proposes the best scheme across combinations of distance measures and clustering methods. The exercise does not run it.

### Interpreting the dendrogram

Each leaf of the dendrogram is one township. Similar townships are joined into branches lower down, and the height of a fusion shows how different the two groups are. Two townships can only be compared by the height at which their branches first join. Their positions along the horizontal axis say nothing about similarity.

`rect.hclust()` draws boxes around the six clusters. The notes give `border = 2:5`, which has only four colours, so two clusters would reuse colours. I gave it six.

```{r}
#| fig-width: 12
#| fig-height: 6
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:7)
```

### Visually driven hierarchical clustering

**heatmaply** combines the dendrogram with a heatmap of the variables, so the reason each township sits in its cluster can be read off the colours.

#### Transforming the data frame into a matrix

`heatmaply()` expects a matrix, so I converted the data frame first.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### Plotting an interactive cluster heatmap

```{r}
#| fig-width: 10
#| fig-height: 10
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

Hovering over a cell shows the township, the variable and its min-max value. This heatmap clusters the normalised values, so its six groups are not exactly the same as the ones from `hclust_ward`, which used the raw values. The darkest block, high on television, mobile phones and computers at once, holds the border trade towns Muse, Tachileik, Laukkaing and Mongla together with Taunggyi, Lashio, Kengtung and Pangsang.

### Mapping the clusters

I cut the tree into six clusters with `cutree()`.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

`groups` is a factor with one label per township. To map it, I converted it to a matrix, attached it to `shan_sf` with `cbind()` and renamed the new column to `CLUSTER`.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
#| fig-width: 7
#| fig-height: 7
qtm(shan_sf_cluster, "CLUSTER")
```

```{r}
table(shan_sf_cluster$CLUSTER)
```

The clusters are very fragmented. I counted the number of separate pieces each cluster forms on the map, using the contiguity neighbour list from Part 4.

```{r}
nb_check <- poly2nb(shan_sf)

count_pieces <- function(cl) {
  sapply(sort(unique(cl)), function(g) {
    suppressWarnings(n.comp.nb(subset(nb_check, cl == g)))$nc
  })
}

count_pieces(shan_sf_cluster$CLUSTER)
```

Five of the six clusters are split into four or five separate pieces. Only cluster 6 is one connected block, Mongmao, Pangsang, Hopang, Pangwaun and Mongyang in the Wa area along the Chinese border. This is the main limitation of a non-spatial method. Townships end up together because their ICT profiles match, whether or not they are anywhere near each other.

### A check, clustering the standardised variables

Since the raw scale gives `MPHONE_PR` and `TV_PR` most of the weight, I reran the same Ward clustering on the min-max values to see how much the grouping depends on that choice.

```{r}
groups_std <- cutree(hclust(dist(shan_ict.std), method = "ward.D"), k = 6)

table(raw = groups, standardised = groups_std)
```

The two solutions agree only in part. Cluster 1 from the raw values mostly stays together, but cluster 3 splits three ways and townships from cluster 5 spread across three groups. With the variables on equal terms, radio and land line phones get more say, and several townships change neighbours in the tree. The rest of this page follows the notes and uses the raw values.

::: {.callout-important}
## My takeaway

Ward's method gave the strongest structure, and the gap statistic pointed to six clusters only by a very small margin over four or five. The six clusters make sense in ICT terms, with a border town group high on almost everything and a Wa group with very little radio. On the map they are scattered across Shan State. I also learned that leaving the variables unscaled hands most of the decision to mobile phones and television.
:::

## Part 4, Spatially constrained clustering with SKATER

SKATER (Spatial 'K'luster Analysis by Tree Edge Removal) only allows neighbouring townships to join the same cluster. It links all townships in a minimum spanning tree, where each link is between two neighbours, and then cuts the tree into pieces. Each piece is one cluster and is contiguous by construction.

### Converting to SpatialPolygonsDataFrame

The notes convert `shan_sf` into an sp object because older versions of the SKATER functions only accepted sp.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

::: {.callout-note}
## sf works directly now

Current versions of spdep accept sf objects in `poly2nb()`. I ran `poly2nb()` on both `shan_sf` and `shan_sp` and got identical neighbour lists, so the conversion is no longer needed. I kept it to follow the notes.
:::

### Computing the neighbour list

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

The 55 townships have 264 neighbour links, an average of 4.8 each. Five townships have only two neighbours, regions 3, 5, 7, 9 and 47, which are Ywangan, Mabein, Pekon, Nawnghkio and Pangwaun. Region 8, Lawksawk, has the most with nine.

I plotted the network on top of the township boundaries. The boundaries go first, since the first plot sets the extent and the network would otherwise clip parts of the map. The centroids of the polygons serve as the nodes.

```{r}
#| fig-width: 8
#| fig-height: 8
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))

plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

### Computing the minimum spanning tree

#### Calculating edge costs

`nbcosts()` gives each neighbour link a cost, the euclidean distance between the two townships in the five ICT variables. A low cost means two neighbours have similar profiles.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

I then turned the neighbour list into a weights object with these costs as the weights. Style `"B"` keeps the costs as they are, without row standardisation.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

`S0`, the sum of all weights, is 76,268. Each link appears twice in the list, once in each direction, so this is twice the total cost of the 132 links.

#### Building the minimum spanning tree

`mstree()` finds the set of 54 links that connects all 55 townships with the lowest total cost.

```{r}
set.seed(1234)
shan.mst <- mstree(shan.w)
```

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

The tree has 54 rows, one fewer than the number of townships, since a spanning tree on n nodes always has n minus 1 edges. Each row holds the two townships at the ends of an edge and its cost.

```{r}
head(shan.mst)
```

::: {.callout-note}
## The first rows differ from the notes

My first rows do not match the notes, which start with townships 54 and 48. `mstree()` picks a random starting township when `ini` is not given, and the order of the rows follows the order in which the tree grows. I built the tree from two different seeds and got exactly the same 54 edges with the same total cost of 9,686.6 both times. Only the order changes, and SKATER gives the same clusters as the notes. I added `set.seed()` so the output above stays the same on every render.
:::

```{r}
#| fig-width: 8
#| fig-height: 8
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

Compared with the neighbour network, each township now has only the links needed to keep the tree connected, and there are no loops.

### Computing spatially constrained clusters

`skater()` takes the two node columns of the tree, the data used to update the costs as groups form, and the number of cuts. Six clusters need five cuts.

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

```{r}
str(clust6)
```

`groups` holds the cluster label of each township. `edges.groups` describes each cluster with its nodes, edges and within cluster sum of squares (`ssw`). `ssto` is the total sum of squares, and `ssw` at the top level records the criterion after each cut.

```{r}
ccs6 <- clust6$groups
ccs6
```

```{r}
table(ccs6)
```

The six clusters have 22, 18, 11, 2, 1 and 1 townships. The notes say the first cluster has 12 nodes, but the `node` vector of the first element of `edges.groups` has length 22, matching the table.

```{r}
round(clust6$ssw)
```

The first cut does most of the work. It lowers the criterion from 12,613 to 10,977, and each later cut takes off only between about 390 and 1,020. The last three cuts remove small branches, which is why cluster 4 has two townships and clusters 5 and 6 one each.

```{r}
shan_sf %>%
  st_drop_geometry() %>%
  mutate(SP_CLUSTER = ccs6) %>%
  filter(SP_CLUSTER %in% 4:6) %>%
  select(SP_CLUSTER, TS.x, RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR) %>%
  arrange(SP_CLUSTER)
```

The singletons are extreme townships. Ywangan has the highest radio rate in the state and the sixth lowest television rate. Narphan has the lowest computer rate in the state and the third lowest television rate. The tree can only isolate them by cutting them off, since every possible region around them would contain townships unlike them.

I plotted the pruned tree on the map. The notes give five colours for six groups, so I added a sixth.

```{r}
#| fig-width: 8
#| fig-height: 8
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink", "orange"),
     cex.circles=0.005, 
     add=TRUE)
```

### Visualising the SKATER clusters

```{r}
#| fig-width: 7
#| fig-height: 7
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

For comparison I put the hierarchical clusters and the SKATER clusters side by side.

```{r}
#| fig-width: 12
#| fig-height: 7
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(col_alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(col_alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

Every SKATER cluster is one connected region. Cluster 1 covers the central and southern belt from Hsipaw down through Loilen to Monghsat. Cluster 2 runs along the whole eastern border from Muse and Lashio to Tachileik and Kengtung. Cluster 3 is the western edge from Mongmit down to Pinlaung, including Taunggyi.

```{r}
shan_sf_spatialcluster %>%
  st_drop_geometry() %>%
  group_by(SP_CLUSTER) %>%
  summarise(n = n(),
            mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```

The three large regions have clear profiles. Among them, the eastern border region (cluster 2) has the highest land line, mobile phone and computer rates but the lowest radio rate, 174 per thousand. The western region (cluster 3) has the highest radio rate at 329 and fairly high television. The central belt (cluster 1) is lowest of the three on television, mobile phones and computers.

The price of contiguity shows in the spread within clusters. Cluster 2 joins the border towns of Muse and Tachileik with the Wa townships of Pangsang and Mongmao because they lie on the same border, even though hierarchical clustering put them in different groups. I measured the share of total variance explained by each solution with a small helper.

```{r}
r_squared <- function(cl) {
  x <- as.matrix(shan_ict)
  tss <- sum(sweep(x, 2, colMeans(x))^2)
  wss <- sum(sapply(unique(cl), function(g) {
    xg <- x[cl == g, , drop = FALSE]
    sum(sweep(xg, 2, colMeans(xg))^2)
  }))
  1 - wss / tss
}

c(hclust = r_squared(shan_sf_cluster$CLUSTER),
  skater = r_squared(ccs6))
```

The hierarchical clusters explain 76 percent of the variance in the five variables, and the SKATER clusters only 50 percent.

::: {.callout-important}
## My takeaway

SKATER turned the scattered pieces into three large connected regions, west, centre and eastern border, plus three tiny clusters for townships too unusual to fit anywhere. That contiguity cost a lot of homogeneity, with the explained variance falling from 76 to 50 percent. Asking for six clusters did not give six useful regions here, because three of the five cuts went into trimming outliers.
:::

## Part 5, Spatially constrained clustering with ClustGeo

**ClustGeo** offers a softer constraint. Its `hclustgeo()` function runs a Ward-like hierarchical clustering on two dissimilarity matrices at once. D0 holds the differences in the clustering variables and D1 holds the differences in space. A mixing parameter alpha between 0 and 1 sets the balance. At alpha = 0 only D0 counts, and at alpha = 1 only geography counts. The aim is to find an alpha that makes the clusters more compact on the map without losing too much homogeneity in the variables, and `choicealpha()` helps with that.

### Ward-like hierarchical clustering with ClustGeo

With only D0, `hclustgeo()` is an ordinary Ward clustering. D0 has to be a `dist` object, so I reused `proxmat`.

```{r}
#| fig-width: 12
#| fig-height: 6
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:7)
```

#### Mapping the clusters

```{r}
#| fig-width: 7
#| fig-height: 7
groups <- as.factor(cutree(nongeo_cluster, k=6))

shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)

qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

This map is close to the `hclust()` map from Part 3 but not the same.

```{r}
table(hclustgeo = shan_sf_ngeo_cluster$CLUSTER, 
      hclust_ward.D = shan_sf_cluster$CLUSTER)
```

Clusters 3, 4 and 6 are identical, and eight townships move between clusters 1, 2 and 5.

::: {.callout-note}
## hclustgeo() matches ward.D2

`hclustgeo()` applies Ward's criterion to squared distances, which is what `hclust()` calls ward.D2. I ran `hclust(proxmat, method = "ward.D2")` and got exactly the same six clusters as `hclustgeo(proxmat)`. The difference above comes from Part 3 using ward.D. The `agnes()` coefficient of 0.943 also belongs to ward.D2, since `agnes(method = "ward")` gives the same tree as ward.D2.
:::

```{r}
count_pieces(shan_sf_ngeo_cluster$CLUSTER)
```

Without the spatial term these clusters are just as fragmented as before, with only cluster 6 in one piece.

### Spatially constrained hierarchical clustering

D1 is the distance between townships. `st_distance()` returns a matrix of distances between polygons, and `as.dist()` turns it into a `dist` object.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

`st_distance()` measures from boundary to boundary, so the 132 pairs of townships that share a border all have a distance of 0. The longest distance is about 444 km.

`choicealpha()` runs the clustering for alpha from 0 to 1 in steps of 0.1 and reports how homogeneous the six clusters are in each space.

```{r}
#| fig-width: 8
#| fig-height: 5
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

```{r}
round(cbind(cr$Q, cr$Qnorm), 3)
```

Q0 measures homogeneity in the ICT variables and Q1 in space, and the normalised versions are scaled to 1 at their best. At alpha = 0, Q0 is 0.757 and Q1 only 0.131. Moving to alpha = 0.2 raises Q1 to 0.626 while Q0 falls to 0.647, or 85 percent of its maximum. Beyond 0.2 the trade gets worse. Alpha = 0.4 gains another 0.2 in normalised Q1 but loses 0.16 in normalised Q0. Following the notes, I use alpha = 0.2.

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
#| fig-width: 7
#| fig-height: 7
qtm(shan_sf_Gcluster, "CLUSTER")
```

```{r}
table(shan_sf_Gcluster$CLUSTER)
count_pieces(shan_sf_Gcluster$CLUSTER)
```

The six clusters are more balanced in size than SKATER's, from 7 to 12 townships, and more compact than the non-spatial ones. Clusters 2 and 5 are fully connected, and clusters 1 and 6 form two pieces each. ClustGeo does not force contiguity, though. Cluster 3 still has five pieces and cluster 4 is spread over seven.

```{r}
shan_sf_Gcluster %>%
  st_drop_geometry() %>%
  filter(CLUSTER == "4") %>%
  pull(TS.x)
```

Cluster 4 holds the main towns of the state, Taunggyi, Lashio, Kyaukme, Muse, Namhkan, Laukkaing, Tachileik, Hseni and Mabein. Their ICT profiles are so close to each other, and so far from the townships around them, that a spatial weight of 0.2 is not enough to split the group by location. For a planning map this group reads as an urban type, and it would need a separate treatment from the regions around it.

I put the four solutions side by side and compared how much variance each explains.

```{r}
#| fig-width: 12
#| fig-height: 12
tmap_arrange(qtm(shan_sf_cluster, "CLUSTER") + tm_title("hclust, ward.D"),
             qtm(shan_sf_ngeo_cluster, "CLUSTER") + tm_title("ClustGeo, alpha = 0"),
             qtm(shan_sf_spatialcluster, "SP_CLUSTER") + tm_title("SKATER"),
             qtm(shan_sf_Gcluster, "CLUSTER") + tm_title("ClustGeo, alpha = 0.2"),
             ncol = 2)
```

```{r}
tibble(method = c("hclust ward.D", "ClustGeo alpha = 0", "ClustGeo alpha = 0.2", "SKATER"),
       r_squared = c(r_squared(shan_sf_cluster$CLUSTER),
                     r_squared(shan_sf_ngeo_cluster$CLUSTER),
                     r_squared(shan_sf_Gcluster$CLUSTER),
                     r_squared(shan_sf_spatialcluster$SP_CLUSTER)),
       pieces_on_map = c(sum(count_pieces(shan_sf_cluster$CLUSTER)),
                         sum(count_pieces(shan_sf_ngeo_cluster$CLUSTER)),
                         sum(count_pieces(shan_sf_Gcluster$CLUSTER)),
                         sum(count_pieces(shan_sf_spatialcluster$SP_CLUSTER)))) %>%
  mutate(r_squared = round(r_squared, 3))
```

The two non-spatial solutions explain about 76 percent of the variance and break into 24 and 28 pieces on the map. ClustGeo with alpha = 0.2 explains 65 percent with 18 pieces, and SKATER explains 50 percent with exactly 6. ClustGeo sits between the two extremes, which is what the mixing parameter is for.

::: {.callout-important}
## My takeaway

ClustGeo gave me a dial between homogeneity and compactness, and `choicealpha()` made the trade visible. At alpha = 0.2 most regions became compact while keeping 85 percent of the homogeneity. It also showed something SKATER hides, a group of main towns spread across the state that are more like each other than like their neighbours.
:::

## Part 6, Visual interpretation of clusters

The notes profile the clusters from `shan_sf_ngeo_cluster`, the ClustGeo solution with alpha = 0, so I used the same object.

### Visualising one clustering variable

```{r}
#| fig-width: 8
#| fig-height: 5
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

Cluster 3 has the highest radio ownership, with a median of 278 households per thousand. Cluster 2 follows at 227 and cluster 1 at 205. Cluster 4 is lower at 155, and clusters 5 and 6 are lowest, both with a median of 97.

::: {.callout-warning}
## The ranking in the notes

The notes rank the clusters 3, 2, 1, 4, 6 and 5 by mean radio ownership. The means in the summary table at the end of the notes, which I reproduce below, give 3, 2, 1, 4, 5 and 6, with cluster 5 at 124 and cluster 6 at 98.6. The boxplot shows medians, and on medians clusters 5 and 6 are tied.
:::

### Multivariate visualisation

A parallel coordinate plot shows all five variables at once, with one line per township. `ggparcoord()` from **GGally** draws it, and `facet_grid()` gives each cluster its own panel. Columns 17 to 21 of `shan_sf_ngeo_cluster` are the five clustering variables.

```{r}
#| fig-width: 14
#| fig-height: 6
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

Cluster 4 stands out with the highest television and mobile phone ownership, and it also leads on land lines and computers. These are the six border and trade towns, Muse, Tachileik, Laukkaing, Mongla, Namhkan and Taunggyi.

Cluster 5 has the lowest television, mobile phone and computer rates of all six clusters. The notes say it is lowest on all five items, but cluster 6 has less radio and cluster 2 has fewer land lines.

`scale = "globalminmax"` draws the raw values on one shared axis. `ggparcoord()` offers five other options. `std` subtracts the mean and divides by the standard deviation, and `robust` subtracts the median and divides by the median absolute deviation. `uniminmax` rescales each variable to run from 0 to 1. `center` applies uniminmax and then centres each variable on a summary value set by `scaleSummary`, and `centerObs` does the same around one chosen observation set by `centerObsID`.

With `globalminmax`, television and mobile phones fill most of the height and computers are squashed along the bottom. I also drew the same plot with `uniminmax`, which puts every variable on the same 0 to 1 range.

```{r}
#| fig-width: 14
#| fig-height: 6
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plots of ICT Variables by Cluster, uniminmax scale") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

On this scale the computer rate of cluster 4 becomes visible, and so does the land line profile of cluster 6, which sits fairly high even though its radio rate is the lowest. No scale is best for every purpose, and switching between them helped me read the clusters.

### Summary statistics

I computed the mean of each clustering variable by cluster to back up the plots.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```

The six clusters can be read as six ICT profiles. Cluster 4 is the connected urban group, with 744 televisions and 651 mobile phones per thousand households and a computer rate of 65.5. Cluster 3 has the most radios (300) and fairly high television (611). Cluster 6, the Wa townships, has the fewest radios (98.6) but relatively high land line and mobile phone ownership. Cluster 5 is the least connected group, with 224 televisions, 132 mobile phones and 6.7 computers per thousand. Clusters 1 and 2 sit in between, with cluster 2 especially low on land lines (23.9).

## Wrapping up

The five ICT rates group Shan State's 55 townships into six profiles that range from a connected urban group to a group with very low ownership of every item. Without a spatial constraint these profiles are scattered across the state in 24 to 28 separate pieces. SKATER produced three connected regions, west, centre and eastern border, plus three outlier clusters, at the cost of explaining only 50 percent of the variance. ClustGeo with alpha = 0.2 kept 65 percent, made most clusters compact, and left the main towns together as one scattered urban group.
